---
title: Setting Up renv
author: Sean Warlick
date: '2020-04-26'
slug: setting-up-renv
categories:
  - R
  - Packages
tags:
  - Reproducible Research
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
output: blogdown::html_page
---

After switching to Ubuntu as my operating system I needed to re-install all of the R packages I needed to work on my website. These included both the packages needed by **blogdown** and packages I used in specific articles. I didn't know what I was missing until I ran into an error. Installing the missing software packages was time consuming and tedious. 

Not only was the process frustrating, recreating my development environment piece by piece did not follow best practices for reproducibility. One of the keys to reproduceibilty is the ability to easily and precisely replicate the computing environment, including all the supporting packages needed.[^1] In support of this goal a number of tools exist to manage dependencies. Most python users are familiar with conda environments, which makes managing project dependencies very easy. For managing package dependencies in R **packrat** has been around for a while, but it has never seemed easy to use. A more recent addition is [**renv**](https://rstudio.github.io/renv/index.html), which I'm going to setup to manage my the dependencies for my website.

### Installing and Setting Up renv
Installing **renv** and setting it up is fairly straight forward. 

```{r eval=FALSE}
install.packages('renv')
renv::init()
```

When `renv::init()` is called three of things happen. First, **renv** searches the project files for all the packages used within the project and copies them to a project specific library. Second, it captures the state of the library in *renv.lock* file. Finally, it builds the infrastructure needed to activate the project specific library each time you launch the project in RStudio. If we look at the figures below we can see the file structure before and after running `renv::init()`. In the after figure we have a new file *renv.lock* and a new directory `renv/`. Inside `renv/` is activate.R which activates the project library and another directory `renv/library/` which stores the project specific library.

```{r images, echo=FALSE}
before_path <- '/post/2020-04-26-setting-up-renv_files/file_structure_before.png'
after_path <- "/post/2020-04-26-setting-up-renv_files/post_init.png"
after_caption = "File Structure After Running `renv::init()`. Notice the addition of renv.lock and renv/"
blogdown::shortcode("figure",
                    src = before_path,
                    caption = "File Structure Before Running `renv::init()`")
blogdown::shortcode("figure", 
                    src = after_path,
                    caption = after_caption)
```

### Installing Packages Manually & Updating Lockfile 
Now that I have project specific library set up I can largely go about my normal workflow, installing packages as needed. For example when I went to preview my updates I discovered that `renv::init()` missed **sourcetools** and **miniUI** because of the limitations with `renv::dependencies()`[^2]. I can add these to my project library using `install.packages()` or `renv::install()`

```{r eval=FALSE}
install.packages('sourcetools')
install.packages('miniUI')
```

Unfortunately installing these packages doesn't automatically update the lockfile. In fact with **renv**'s default settings for the snapshot type, it thinks the lockfile is still up to date. Fixing this requires changing the settings to capture the entire project library[^3]. After changing the setting, I can simply run `renv::snapshot()` any time we install new packages to our project requirements current. I can also run `revn::status()` to check if the lockfile is synced with the project library. 


```{r, collapse=TRUE}
renv::snapshot() # Nothing has changed becasue of the default setting
```

```{r eval=FALSE}
# Change the snapshot setting
renv::settings$snapshot.type("simple")

renv::snapshot()

## The following package(s) will be updated in the lockfile:
## 
## # CRAN ===============================
## - KernSmooth    [* -> 2.23-16]
## - boot          [* -> 1.3-24]
## - class         [* -> 7.3-16]
## - cluster       [* -> 2.1.0]
## - codetools     [* -> 0.2-16]
## - fastmap       [* -> 1.0.1]
## - foreign       [* -> 0.8-75]
## - miniUI        [* -> 0.1.1.1]
## - nnet          [* -> 7.3-13]
## - rpart         [* -> 4.1-15]
## - shiny         [* -> 1.4.0.2]
## - sourcetools   [* -> 0.1.7]
## - spatial       [* -> 7.3-11]
## - survival      [* -> 3.1-11]
## - xtable        [* -> 1.8-4]
## 
## * Lockfile written to '~/Documents/website-source/renv.lock'.
```

### renv and Version Control
Now that I have all the software dependencies documented, I need get everything under version control to make it truly reproducible. Naturally, I don't want to commit my entire project library. I just want to commit three files 1) renv.lock, 2) .Rprofile, and 3) renv/activate.R. With these three files present in the repository all I need to do is call to recreate my environment `renv::restore()`

**renv** is designed to integrate well with version control.  To see past versions of the lockfile committed to Git I can call `renv::history()`. Using `renv::revert(commit = commit)`, I can revert to a previous version of the file.

### Conclusions
Creating a reproducible development environment has been greatly simplified by **renv**.  With only two functions, `renv::init()` and `renv::snapshot()` I can keep track of the packages I need to preview and build my website.  If I ever need to recreate my environment, I won't need to install packages one by one as I encounter errors in the build.  All I will need to do is call `renv::restore()`. 


[^1]:The Turing Way Community, Becky Arnold, Louise Bowler, Sarah Gibson, Patricia Herterich, Rosie Higman, … Kirstie Whitaker. (2019, March 25). The Turing Way: A Handbook for Reproducible Data Science (Version v0.0.4). Zenodo. http://doi.org/10.5281/zenodo.3233986;
Gandrud, Chrisopher, *Reproducible Research with R and RStudio
Second Edition* (CRC Press, 2016), 10.

[^2]: See `?renv::dependencies` for details on how **renv** finds packages used in a project
[^3]: See `?renv::snapshot` for details on the snapshot settings. 
